# The basics 

## installing RStudio 

## The RStudio IDE

## R Projects

## Functions and Packages



outline
- install the tidyverse
- the tidyverse is a collection of packages that make data analysis and cleaning and everythingg else a lot easier

```{r eval=FALSE}
install.packages("tidyverse")
```

- a package is a collection of functions that make life easier. functions are bits of code that do things. we will use _a lot_ of functions throughout. we will dive into this in more depth later on. You can usually recognize a function by the parentheses at the end i.e. `function()`
  - think about the anatomy of a function:
    - function name and arguments
    - i will be refering to "function arugments" a TON so please internalize this. 
    - we will review _making_ functions later. But using functions, oof. Main stay
    - arguments are unquoted
- the way we load this package for use with the `library()` function (a little meta, lol)


- this one line of code will install the packages that we need to do this work. Don't be too concerned about _what_ a package is. We will go over this in more detail later. Very simply, a package is a collection of functions that make certain things possible or easier to do in R. We like to use packages because reinventing the wheel is never fun.

```{r, eval = FALSE}
install.packages("tidyverse")
```

