# Broken Windows and the origin of Ecometrics


```{r, include=FALSE}
#source("common.R")
```

To Cover:

- sampson & raudenbush systematic social observation: https://www.journals.uchicago.edu/doi/abs/10.1086/210356
- ecometrics: where was this first coined for the social sciences????
- 

-----

- 1980 "Broken Windows" in the Atlantic
- Posited that the presence of physical disorder led to crime
- this influenced policing policies like stop and frisk
  - led to the infamous implementation of [CompStat](https://nymag.com/intelligencer/2018/03/the-crime-fighting-program-that-changed-new-york-forever.html)
- the simplicity of the theory was appealing to public officials. media lapped it up
- social scientists were quick to shoot down the theory, seemed to not matter
- come 1999 Sampson and Raudenbush seek to test this find no conclusion
- much more work comes from it
- most noteably is the use of ecometrics in assessing this. I say most noteably because of its approach
  - it combines an inductive approach with a deductive one.
  - this is the quintessence of the Boston approach, of course it comes from Dan O'Brien.
    - why?:
      1. big data (actually big-ish which is rare in social science)
      2. The development and use of ecometrics (induction)
      3. deliberately testing an existing theory (deduction)
      4. Absolutely crushing it.
- what is an ecometric?
  - why is it inductive?
    - taking naturally occuring data, and extracting latent (already existing variables)
    - 


> Note: a case study should be recreating these ecometrics

Explore what the data description from BARI looks like

```{r}
library(tidyverse)
ecometrics <- readr::read_csv("data/911/911-ecometrics-2014-19.csv") %>% 
  janitor::clean_names()

glimpse(ecometrics)
```

Check out 

```{r}
raw_911 <- read_csv("data/911/911-raw.csv", n_max = 5000) %>% 
  janitor::clean_names()
```

```{r rows.max=20}
raw_911 %>% 
  filter(str_detect(offense_description, coll("battery", T)))
```




### Resources:

- https://www.annualreviews.org/doi/abs/10.1146/annurev-criminol-011518-024638?journalCode=criminol
- Large-scale data use, ecometrics to assess disorder: https://journals.sagepub.com/doi/abs/10.1177/0022427815577835
