# General data manipulation

You may have heard of the 80/20 rule—or at least one of the many 80/20 rules.
Data scientists are often fond of emphasising that 80% of their work is data cleaning and the other 20% is analysis and modeling
this is all to say that you will find yourself with messy, unsanitized, gross, not fun to look at data most of the time
because of this, it's really important that we have the skills to clean our data 
right now we're going to go over the foundational skills we will learn how to select columns and rows, filter data, and create new columns and variables
we will be using functions from the package dplyr which is imported from the tidyverse

- recall how to load the tidyverse
- we'll read in the `ACS_1317_TRACT.csv` file located in the `data` directory
  - putting this together the file path is `data/ACS_1317_TRACT.csv`.
  - store it in the variable `acs_raw`
  

```{r message=FALSE}
library(tidyverse)
acs_raw <- read_csv("data/ACS_1317_TRACT.csv")

glimpse(acs_raw)
```

This is the dataset from which `acs_edu` was created. We will work to create the same dataset. For a refresher the columns were: "med_house_income, less_than_hs, hs_grad, some_coll, bach, white, and black.

```{r message=FALSE}
glimpse(read_csv("data/acs_edu.csv"))
```


to do so we use the function `select()`. Look at the help documentation by running `?select()` inside of the console. What are the arguments?
- `.data` and `...`. We'll tackle this one by one.
  - `.data`: "A tbl." 
    - read this as "A tibble, or data frame". 
    - for those of you who are curious, `tbl` is the formal object class of a tibble. a tibble is still a data frame
    - this means that the first argument will be passed the `acs_raw` object
- `...`: 

> One or more unquoted expressions separated by commas. You can treat variable names like they are positions, so you can use expressions like x:y to select ranges of variables.

- this one is a little bit more complex. `...` referred to as "dots". This really means we can pass any number of other arguments to `select()` here the `...` will be ways of referring to the columns we want. and there are a number of ways we can referrer to which columns we want we can:
  1. write the name (unquoted) of the columns—i.e. i want the column `col_x` would be `select(.data, col_x)`
  2. write the position of the column—i.e. i want the first column would be `select(.data, 1)`
  3. There are also a number of _select helpers_ of which we will go over a few
- one behavior that you should notice is that select will _always_ return a tibble, even if we don't sepcify any columns to select
- try using the select function on `acs_raw` without passing any column specifications to `...`

```{r}
select(acs_raw)
```

- select the median house hold income column (`med_house_income`)

```{r}
select(acs_raw, med_house_income)
```


- as the documentation notes we can select a range of columns using `col_x:col_y`
- select the columns relating to education from less than high school to doctoral

```{r}
select(acs_raw, less_than_hs:doc)
```

- though you will likely not use it too often, it's still important to be comfortable with column index (position) selecting
- select the 1st, 5th, and 10th columns

```{r}
select(acs_raw, 1, 5, 10)
```



- select
  - by position or name
- slice
- filter
  - logical operators
- mutate

_i need to introduce measures of central tendency and summar stats in between this_

- groups
- summarizing
