# Visualizing beyond 2-dimensions


```{r, include=FALSE, message=FALSE}
source("common.R")
library(tidyverse)

acs_messy <- read_csv("data/ACS_1317_TRACT.csv")

acs <- acs_messy %>% 
  separate(name, sep = ", ", into = c("tract", "county", "state")) %>% 
  mutate(tract = str_remove(tract, "Census Tract ")) %>% 
  na.omit()
```

To cover:

- coloring / filling
- shaping
- sizing
- faceting

## Expanding bivariate visualizations to trivariate & other tri-variate



- we can visualize other vcariables by setting further aesthetics.
  - can set the color or fill, size, and shape
- we alreay did this previously when we set the color, let's do that here. 
  - lets see how commuting by walking changes with the family house and under 18 pop
    - set the color argument of the `aes()` function as `color = by_walk`
      - it's important you do this within the aesthetics function 
    
```{r}
ggplot(acs, aes(fam_house_per, age_u18, color = by_auto)) +
  geom_point()
```

- we can add size to this as well by setting the `size` aesthetic
  - lets see if the more female headed house holds there are affects commuting by car as minors increases

```{r}
ggplot(acs, aes(fam_house_per, age_u18, color = by_auto, size = fem_head_per)) +
  geom_point(alpha = .2)
```

- from this chart we can see quite a few things:
  - as `fam_house_per` increases so does the under 18 pop,
  - as both `age_u18` and `fam_house_per` increase so does the rate of communiting by car
  - as both `age_u18` and `fam_house_per` so does female headed houses, but to a lesser degree
  - this gives us a good idea of some relationships that we can test with our data at a later point

```{r}
minors_lm <- lm(age_u18 ~ fam_house_per + by_auto + fem_head_per, data = acs)

huxtable::huxreg(minors_lm)
```



Trivariate:

- grouped / stacked bar charts
- heatmaps 


Color Ramps:

- diverging when there is a true middle
- dark is low bright is high


- most data analyses start with a visualization. 
- the data we have will dictate the type of visualizations we create
- there are many many different ways in which data can be represented
- generally these can be bucketed into a few major categories
  - numeric 
    - integer
    - double
  - character 
    - think groups, factors, nominal, anything that doesn't have a numeric value that makes sense to count, aggregate, etc.
  - time / order 