# Visualizing beyond 2-dimensions

```{r, include=FALSE, message=FALSE}
source("common.R")
library(tidyverse)

acs_raw <- read_csv("data/ACS_1317_TRACT.csv")

commute <- select(acs_raw,
       county,
       hs_grad, bach, master,
       starts_with("commute"),
       starts_with("by"),
       med_house_income) %>% 
  filter(county %in% c("SUFFOLK", "NORFOLK", "MIDDLESEX")) %>% 
  mutate(hh_inc_quin = ntile(med_house_income, 5),
         edu_attain = bach + master) %>% 
  select(-bach, -master)
```


Over the duration of the last three chapters we have cultivated a fundamental understanding of the grammar of graphics and have discussed how to craft univariate visualizations and explore bivariate relationships. In those graphics we have not gone beyond two-dimensions. We only utilized two aesthetics to map. Howver, there is many more that we can incorporate into our visualizations which will in turn enable us to explore three or for variables at once.

> Note that these will not be 3D, but rather visualize three variables. 

To improve our graphics we will utilize the color, shape and size aesthetics as well as faceting. Of course, this begs the question of which aesthetic do I choose? Well, that depends upon what type of data you will be visualizing. Each aesthetic serves a different purposes and can be used for a different type of variable.

In general we can use the below mappings: 

- color -> continuous or discrete
- size -> continuous
- shape -> discrete

- Let's first look at the use of color. 
- lets first look at the use of color for discrete measures
- look at the ecometrics for social disorder discussed earlier and developed in o'brien 2015


- we have year, measure, and n, lets view year on the x and n on the y
  - we now want to know these totals by measure
  - set the `fill` to the `measure`. fill is used for polygons (a col is a rectangle that needs to be filled)
  - by default this shows us a stacked bar chart which is useful
- another way to do this is to change the position of these bars by setting the position argument to `"dodge"` this enables us to compare the heights of the bars more easily
- also, we should always consider creating small multiples, or facetting, rather than adding more to our plots. We want the data to be consumable. Don't feel forced to put everything on to one graphic

- Continuous use of color.
  - there are two general ways in which continuous variables will be mapped to color. this is if there is a low value and a high value.
  - secondly, we can think of a divergent color palette. This is used when there is a known middle value
    - a common mistake is using a divergent palette when there is no natural middle. be wary of this. 
    
 
```{r message = FALSE}
commute <- read_csv("data/acs_edu.csv")

commute
```
 
 
 



```{r}
ecometrics <- read_csv("data/ecometrics.csv")

# viewing by fill
ggplot(ecometrics, aes(year, n, fill = measure)) +
  geom_col()


# viewing by fill
ggplot(ecometrics, aes(year, n, fill = measure)) +
  geom_col()

# viewing by fill and dodging
ggplot(ecometrics, aes(year, n, fill = measure)) +
  geom_col(position = "dodge")

# faceting 
ggplot(ecometrics, aes(year, n)) +
  geom_col() + 
  facet_wrap("measure", nrow = 1)


```


we can utilize color, size, and shape to tell a better story with our graphics. 
Each aesthetic can serve a different purpose / used for a different kind of variable

others:
- linetype 

To cover:

- coloring / filling
- shaping
- sizing
- faceting

## Expanding bivariate visualizations to trivariate & other tri-variate


- we can visualize other vcariables by setting further aesthetics.
  - can set the color or fill, size, and shape
- we alreay did this previously when we set the color, let's do that here. 
  - lets see how commuting by walking changes with the family house and under 18 pop
    - set the color argument of the `aes()` function as `color = by_walk`
      - it's important you do this within the aesthetics function 
    
```{r}
ggplot(acs_raw, aes(fam_house_per, age_u18, color = by_auto)) +
  geom_point()
```

- we can add size to this as well by setting the `size` aesthetic
  - lets see if the more female headed house holds there are affects commuting by car as minors increases

```{r}
ggplot(acs_raw, aes(fam_house_per, age_u18, color = by_auto, size = fem_head_per)) +
  geom_point(alpha = .2)
```

- from this chart we can see quite a few things:
  - as `fam_house_per` increases so does the under 18 pop,
  - as both `age_u18` and `fam_house_per` increase so does the rate of communiting by car
  - as both `age_u18` and `fam_house_per` so does female headed houses, but to a lesser degree
  - this gives us a good idea of some relationships that we can test with our data at a later point

```{r}
minors_lm <- lm(age_u18 ~ fam_house_per + by_auto + fem_head_per, data = acs_raw)

huxtable::huxreg(minors_lm)
```



Trivariate:

- grouped / stacked bar charts
- heatmaps 


Color Ramps:

- diverging when there is a true middle
- dark is low bright is high


- most data analyses start with a visualization. 
- the data we have will dictate the type of visualizations we create
- there are many many different ways in which data can be represented
- generally these can be bucketed into a few major categories
  - numeric 
    - integer
    - double
  - character 
    - think groups, factors, nominal, anything that doesn't have a numeric value that makes sense to count, aggregate, etc.
  - time / order 
  
-------


Notes:

https://extremepresentation.typepad.com/files/choosing-a-good-chart-09.pdf