# Creating Ecometrics

```{r, include=FALSE}
source("common.R")
```


> Note: a case study should be recreating these ecometrics

Explore what the data description from BARI looks like

```{r}
# library(tidyverse)
# ecometrics <- readr::read_csv("data/911/911-ecometrics-2014-19.csv") %>% 
#   janitor::clean_names()
# 
# glimpse(ecometrics)
```

Check out the offense descriptions

```{r}
# raw_911 <- read_csv("data/911/911-raw.csv") %>% 
#   janitor::clean_names()
```

```{r}
# raw_911 %>% 
#   count(offense_code_group)
```

- There are fewer types of code groups. 
- They are rather informative.
- 33k missing observations though
- is the same amount of missingness present in the descriptions?

```{r}
# raw_911 %>% 
#   count(offense_description)
```

- no missingness
- more types
- what do we do?
  - choose one of these?
  - we can actually use both
  - filter data set to exclude NAs from `offense_code_group`.
  - `anti_join()` that dataset from raw
  - now use the raw_911 remainders and hand code that.
- reminder: we cannot automate everything. It requires long hours, dedication, and freakin' grit.
